
```{r setup, include=FALSE,error=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=FALSE,message=FALSE)
```


## Introduction

## Prerequsites 

```{r}
library(tidyverse)
library(patchwork)
library(sf)
library(terra)
library(stars)
library(ggspatial)
```

Load in the CPAD_2023a_SuperUnits.shp and the ghm.tif files. ghm.tif Transform both to EPSG:4326.

```{r}
cpad_super <- st_read(here::here("week-10", "discussion_data", "CPAD_2023a_SuperUnits.shp"), quiet = TRUE) %>%
  sf::st_transform("EPSG:4326") %>%
  janitor::clean_names() %>%
  mutate(ID = row_number())

ghm <- rast(here::here("week-10", "discussion_data", "gHM_masked.tif")) %>%
  project("EPSG:4326")
```

## Exercises

1. Let’s make nice plots of the California Protected areas by access level

```{r}
ggplot(data = cpad_super) +
  geom_sf(aes(color = access_typ, fill = access_typ)) +
  theme_linedraw() +
  labs(color = "Access Type", fill = "") +
  annotation_scale(plot_unit = "km") +
  annotation_north_arrow(location = "tr") +
  coord_sf() +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  facet_wrap(~access_typ)
```

```{r}
plot(ghm)
```

2. Let's try plotting the ghm layers nicely too!

```{r}
ggplot() +
  geom_stars(data = st_as_stars(ghm)) +
  coord_equal() + 
  scale_fill_viridis_c() +
  theme_bw()
```

3. Create a function to take 2 data sets (1 polygon and 1 raster) and create a boxplot of the values based on a specific layer

```{r}
summary_boxplot = function(polygon, raster, my_layer, my_label) {
  
  # rasterize polygon by layer
id_rast <- rasterize(polygon, raster, field = "suid_nma")
  
  #do mean zonal statistics
zonal_layer <<- zonal(raster, id_rast, run = "mean", na.rm = TRUE)
 
 #join with polygon database
poly_join <<- full_join(polygon, zonal_layer) %>%
  select(suid_nma, gHM, paste(my_layer))
  
  #create boxplot based on your layer
p1 <- ggplot(poly_join) +
  geom_boxplot(aes(gHM, .data[[my_layer]])) +
  theme_linedraw() +
  labs(x = "Human Modification Index", y = my_label)

return(p1)
}

summary_boxplot(cpad_super, ghm, "access_typ", "Access Type")
```

4. Let's select some layers and use our new function!

```{r}

```

```{r}

```

```{r}

```
